---
title: "Part 1: Fundamentals of data import, structuring, and plotting in tidyverse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{group-01}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ubcBIOL548L)

library(gapminder)
library(ggplot2)
library(tidyr)
library(dplyr)
library(tibble)
library(readr)
library(readxl)


```


```{r, include = FALSE}

  file.copy(system.file("extdata", "gapminder.csv",
              package = "ubcBIOL548L"),
            "./gapminder.csv")
  file.copy(system.file("extdata", "gapminder.xlsx",
                        package = "ubcBIOL548L"),
            "./gapminder.xlsx")

```

## Introduction

This is an R workshop created for a graduate data visualization course at the University of British Columbia by Vikram Baliga, Andrea Gaede and Shreeram Senthivasan.  

There are three groups with increasing levels of difficulty. Each group has a set of learning objectives.

## Group 1: Load a clean data file into R and make boxplots and scatterplots

### Learning Objectives:
1. Determine how to import and control data stored in different filetypes
2. Understand the best practices for structuring data in *tidyverse/ggplot*
3. Construct basic plots using *ggplot*

### Lesson:  
#### Importing data:
**Import data using base R command, and give it the name `my_data`**
```{r}
my_data <- read.csv("gapminder.csv")
# In practise, read_csv() is often better

```

**Take a look at your data**
```{r eval=FALSE}
# Same as print(my_data)
my_data
```

```{r}
# Structure
str(my_data)
# Summarize columns
summary(my_data)
# Get column names (good for wide datasets)
names(my_data)
# Get first 6 lines
head(my_data)
# Get last 6 lines
tail(my_data)
```

**Arguments can be added to a function using commas**  
**Note:** arguments with the default setting are hidden, unless specified. Here n changes the default from 6 to 10 lines
```{r}
head(my_data, n = 10)
```

**The helpfile lists what arguments are available**
```{r, eval=FALSE}
?head
```


**A better import option using Tidyverse**
```{r}
my_data <- read_csv("gapminder.csv")
# Cleaner import and print with read_csv, don't need head()
my_data
# Note that words read in as `chr` not `factors`, this is good!
str(my_data)
# But underlying data is the same
summary(my_data)
```
**Other formats for import**
```{r}
my_data <- read_delim("gapminder.csv", ',')
# Looks like a weird error
my_data <- read_excel("gapminder.xlsx")
```

**Ways to clean up your data during import**
```{r}
# Inspect with head, or excel. We see two junk rows:
head(my_data)

# This can be solved by adding an argument `skip` is the number of rows to skip
my_data <- read_excel("gapminder.xlsx",
                   skip = 2)

my_data <- read_csv("gapminder.csv",col_names = FALSE)
# Setting `col_names` to false made the column headers row one and added dummy column names
my_data
```

```{r}
# We're now going to import the gapminder dataset using the preferred read_csv() function
my_data <- read_csv("gapminder.csv",col_names = TRUE)
```
```{r}
# This looks correct. Note: TRUE is the default so was not needed above
my_data
```

#### Plotting:  
**Make a histogram**
```{r}
# This command makes a histogram of the `lifeExp` column of the `my_data` dataset
qplot(x = lifeExp, data = my_data)
```
```{r}
# The same function here makes a scatter plot
qplot(x = gdpPercap, y = lifeExp, data = my_data)
```
```{r}
# The same function here makes a dot plot because the x axis is categorical
qplot(x = continent, y = lifeExp, data = my_data)
```

**How can the same function make three different classes of plots?**  
One of the hidden arguments is `geom` which specifies the type of plot. The default is `auto` which leads to a guess of the plot type based on the
data type(s) in the column(s) you specify.  

Type `?qplot` in the console to read the `qplot` documentation

**Now let's specify the type of plot explicitly**
```{r}
qplot(x = lifeExp, data = my_data, geom = 'histogram')
qplot(x = gdpPercap, y = lifeExp, data = my_data, geom = 'point')
```
```{r}
# Note that we are now specifying boxplot instead of point plot
qplot(x = continent, y = lifeExp, data = my_data, geom = 'boxplot')
```

**Now let's change the number of bins in a histogram and make the plot prettier**
```{r}
# The hidden argument `bins` has a default valute of 30
qplot(x = lifeExp, data = my_data, geom = 'histogram')
```
```{r}
# This changes the number of bins to 10
qplot(x = lifeExp, bins = 10, data = my_data, geom = 'histogram')
```
```{r}
# Alternatively you can choose the width you want the bins to have
qplot(x = lifeExp, binwidth = 5, data = my_data, geom = 'histogram')
```
```{r}
# Let's add a title
qplot(x = lifeExp, binwidth = 5, main = "Histogram of life expectancy", data = my_data, geom = 'histogram')
```
```{r}
# Let's add an x axis label
qplot(x = lifeExp, binwidth = 5, main = "Histogram of life expectancy", xlab = "Life expectancy (years)", data = my_data, geom = 'histogram')
```
```{r}
# Let's add a y axis label
qplot(x = lifeExp, binwidth = 5, main = "Histogram of life expectancy", xlab = "Life expectancy (years)", ylab = "Count", data = my_data, geom = 'histogram')
```
```{r}
# This format is easier to read, but otherwise exactly the same.
# The convention is to break lines after commas.
qplot(x = lifeExp,
      binwidth = 5,
      main = "Histogram of life expectancy",
      xlab = "Life expectancy (years)",
      ylab = "Count",
      data = my_data,
      geom = 'histogram')
```

**Let's apply a log scale and add a trendline to a scatter plot**
```{r, warning=FALSE}
# Note that x axis is compressed
qplot(x = gdpPercap, y = lifeExp, data = my_data, geom = 'point')
# Here the x axis is log transformed
qplot(x = gdpPercap, 
      y = lifeExp, 
      log = 'x',
      data = my_data, 
      geom = 'point')

# Let's add a trendline to the data as well.
# The linear regression model (`lm`) will be added on top of our previous plot
qplot(x = gdpPercap,
      y = lifeExp,
      log = 'x',
      main = "Scatter plot of life expectancy versus GDP per capita",
      xlab = "Log-transformed GDP per capita ($)",
      ylab = "Life expectancy (years)",
      data = my_data,
      # The following line adds a `smooth` trendline
      # We want our regression to be a linear model, or `lm`
      method = 'lm',
      # the `c()` function allows us to pass multiple variables
      # to the `geom` argument
      geom = c('point','smooth'))
```

**Finally let's make a boxplot**
```{r}
qplot(x = continent,
      y = lifeExp,
      main = "Boxplot of life expectancy by continent",
      xlab = "Continent",
      ylab = "Life expectancy (years)",
      data = my_data,
      geom = 'boxplot')
```
